Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-18 E-Hypergraphs-LICS24.tex
Soundness proof update, citation in conclusion fix

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f = [f_1,f_2]: n'' \to \mathcal H$ in the diagram above is defined as in Definition \ref{def:boundary_new} of boundary complement,  and $f_2$ as the restiction of the composite $i'' \to \mathcal R \to \mathcal H$ and,  and similarly for the map $g: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

@@

In the future,  we intend to investigate a number of natural extensions to mathematical treatment of (acyclic) e-graphs developed here.
Some of these are: support for `functorial boxes'~\cite{mellies} (in addition to e-boxes), which have a variety of applications, including the Cartesian-closure required for the implementation of functional programming languages~\cite{GhicaZanasiTutorial}, as sketched in Fig. \ref{fig:app1}~(a); the `spiders' used by the ZX-calculus to model quantum circuits~\cite{coecke_interacting_2011,ZX}, as sketched in Fig. \ref{fig:app1}~(b); and  traces~\cite{Joyal, Hasegawa},  which can be used to express feedback in categorical models of digital circuits~\cite{GhicaJung, GhicaKaye},  and, indeed cycles in conventional e-graphs to encode infinite equivalence classes, as sketched in Fig.~\ref{fig:app2}. All these application domains stand to benefit from the use of optimisation techniques, which in turn can take advantage of the appropriate e-graph technique.

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V};\iota_{V_{X}} (v_i)) = \consistency(f_{V};\iota_{V_{X}}(v_j))$ and $\consistency(g_{V};\iota_{V_{Y}}(v_i)) = \consistency(g_{V};\iota_{V_{Y}}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{c,X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
        \\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

\begin{proposition}
\label{prop:complement_existence}
    Consider a pushout square in $\catname{EHyp_{\Sigma}}$ below
    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJYIl0sWzIsMCwiWiJdLFsyLDIsIkMiXSxbMCwyLCJQXntcXHVyY29ybmVyfSJdLFsyLDUsIlxcYnVsbGV0Il0sWzEsMCwiaSIsMl0sWzEsMl0sWzIsM10sWzAsMywibSIsMl1d
% \[\begin{tikzcd}
% 	X && Z \\
% 	\\
% 	{P^{\urcorner}} && C \\
% 	\\
% 	\\
% 	&& \bullet
% 	\arrow["i"', from=1-3, to=1-1]
% 	\arrow[from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% 	\arrow["m"', from=1-1, to=3-1]
% \end{tikzcd}\]
\[
    \tikzfig{appendix/pushout_square}
\]
    Such that $m,f$ are monos and $\mathcal{K}$ is discrete.
    Then, pushout complement $\mathcal{K} \to \mathcal{L}^{\bot} \to \mathcal{G}$ exists if
    \begin{enumerate}
        \item $[m(v_i)) = [m(v_j))$ for all $v_i,v_j$ that have a pre-image in $V_{\mathcal{K}}$.
        \item $\consistency(m_{V};\iota_{V_{\mathcal{G}}}(v_i)) = \consistency(m_{V};\iota_{V_{\mathcal{G}}}(v_j))$ for all $v_i,v_j$ that have a pre-image in $V_{\mathcal{K}}$.
        \item The no-dangling condition holds.
    \end{enumerate}
\end{proposition}
\begin{proof}
    By requiring that $m$ is mono makes the no-identification condition satisfied on the nose.
    Then, no-dangling and no-identification conditions are necessary for pushout complement to exists~\cite{dpo}.
    Points (1)-(2) above guarantee that after removing from $\mathcal{G}$ everything in $\mathcal{L}$ that has no pre-image in $\mathcal{K}$ the arrow $\mathcal{K} \to \mathcal{L}^{\bot}$ will satisfy the assumptions~\ref{pushout:assumptions}.
\end{proof}

@@

@@

\begin{proposition}[Proposition~\ref{prop:soundness} restatement]
The category $\WellTypedMdaEcospans/{\mathcal{S}}$ is a semilattice-enriched PROP. 
\end{proposition}
\begin{proof}
    To prove this statement we essentially need to show that the functor $\llbracket - \rrbracket$ is well-defined, \textit{i.e.}, if $f = g$ in $\textsf{PROP}^{+}(\Sigma)$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ or $\llbracket f \rrbracket \Rrightarrow_{\mathcal{S}}^{+,*} \llbracket g \rrbracket$ in $\WellTypedMdaEcospans$.
    $f = g$ implies that there is a sequence of SMC and enrichment equations such that when applied to $f$ it is turned into $g$.
    As $\textsf{PROP}^{+}(\Sigma)$ is quotiented by SMC equations we have the following contexts $\mathcal{C}$ for any term $f$ such that $f = \mathcal{C}[l]$ for some other term $l$.
    \[
        \mathcal{C} := [] ~|~ c_1 + \ldots + \mathcal{C} + \ldots + c_n ~|~ c_1 ; (id \otimes \mathcal{C}); c_2
    \]
    we will proceed by induction on the number of equation applied to make $f$ into $g$.
    \begin{itemize}
        \item Base case. $f = g$ implies that for some equation $l = r$ in Figure~\ref{fig:string-equations}, $f = \mathcal{C}[l]$ and $g = \mathcal{C}[r]$.
              We then note that every subterm of $f$ corresponds to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence so does $l$.
              By computing pushout complement we remove the image of $\llbracket l \rrbracket$ from $\llbracket f \rrbracket$ and replace it with $\llbracket r \rrbracket$ which yields $\llbracket g \rrbracket = \llbracket \mathcal{C} \rrbracket [\llbracket r \rrbracket] = \llbracket \mathcal{C}[r] \rrbracket$ and $\llbracket f \rrbracket \Rrightarrow^{+}_{l = r} \llbracket g \rrbracket$.
        \item Suppose $f = g$ by applying less than $n$ equations implies $\llbracket f \rrbracket \Rrightarrow^{+,*}_{\mathcal{S}} \llbracket g \rrbracket$
        \item Let $f = g$ by applying exactly $n$ equations and then $f = g'$ by applying first $n-1$ equations and $\llbracket f \rrbracket \Rrightarrow^{+,*}_{\mathcal{S}} \llbracket g' \rrbracket$ and $g' = g$ by applying one equation.
              By reusing the base case we have that $\llbracket g' \rrbracket \Rrightarrow^{+}_{l = r} \llbracket g \rrbracket$ and hence $\llbracket f \rrbracket \Rrightarrow^{+,*}_{\mathcal{S}} \llbracket g \rrbracket$.
    \end{itemize}
    The only thing that is left to show that pushout complement and pushout exist for every equation in $\mathcal{S}$.
    Suppose, $f = g$ by applying $l = r$ which is the top left equation from~\ref{fig:string-equations}.
    $\llbracket r \rrbracket$ is then given in Figure~\ref{fig:structural_rule_pushout}.
    \begin{figure*}
        \centering
        \[
            \scalebox{0.45}{
                \tikzfig{combinatorial_semantics/structural_rule_pushout}
            }
        \]
        \caption{The image of the top left equation from Figure~\ref{fig:string-equations} under $\llbracket - \rrbracket$}
        \label{fig:structural_rule_pushout}
    \end{figure*}

    After we identify a convex down-closed image of the left-hand side of this rule within a graph $\llbracket f \rrbracket$ we remove the image by keeping only the top-level verticies.
    Note that because the image of all $a$'s and $b$'s are top-level verticies, we are allowed to pick any monomorphism for our pushout complement.
    And for the same reason the pushout for the right square exists.
    Then we simply glue the right-hand side of the rule into the hole.
    Similar argument applies to other rules as well.
\end{proof}

\begin{proposition}[Proposition~\ref{prop:wnormal_form} restatement]

For each cospan $f$ in ${\WellTypedMdaEcospans}/{\mathcal{S}}$ there is a \textit{weak normal form} such that 
\[
    f = f_1 + \ldots + f_n
\]
such that each $f_i$ contains no hierarchical edges,  and for all $i \neq j$ we have $f_i \neq f_j$.
\end{proposition}    
\begin{proof}
    First, we apply the interpretation of the rules (1)-(4) from Figure~\ref{fig:string-equations} which reduces the number of edges and vertices until there is only one top-level edge which is hierarchical.
    There will be a redex for these rules as long as there are at least two connected edges at least one of which is hierarchical.
    The above steps therefore produce a form like \[f_1 + \ldots + f_n\] where each $f_i$ is also of the form $g_1 + \ldots + g_m$ and so on.
    Then we apply the interpretation of rules (5)-(6) and (8) which reduces the number of hierarchical edges at each step.
    Once again, there is a redex for such a rule as long as there are at least two hierarchical edges.
    This will yield a form $f_1 + \ldots + f_n$ such that each of $f_i$ does not contain any hierarchical edges and for $i \not = j$, $f_i \not = f_j$.
    The latter is guaranteed by rule (8).
\end{proof}

>>>>

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f_1 = [g_1,h_1]: n'' \to \mathcal H$ in the diagram above consists of $g_1$ as defined in Definition \ref{def:boundary_new} of boundary complement,  and $h_1$ which is the restiction of the composite $i'' \to \mathcal R \to \mathcal H$ to $i'' \setminus i$,  and similarly for the map $f_2: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
    Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ if there exists a sequence of such rewrites. 
\end{definition}
The boundary complement conditions, in particular, prohibits finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{combinatorial_semantics/f_plus_g_inline}}
\]
Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

@@

In the future,  we intend to investigate a number of natural extensions to mathematical treatment of (acyclic) e-graphs developed here.
Some of the extensions that can be considered are: support for `functorial boxes'~\cite{mellies_functorial_2006} (in addition to e-boxes), which have a variety of applications, including the Cartesian-closure required for the implementation of functional programming languages~\cite{ghica_hierarchical_2023}, as sketched in Fig. \ref{fig:app1}(a); the `spiders' used by the ZX-calculus to model quantum circuits~\cite{coecke_interacting_2011,ZX}, as sketched in Fig. \ref{fig:app1}(b); or trace~\cite{joyal_geometry_1991, Hasegawa-traced} which can be used to express feedback in categorical models of digital circuits~\cite{ghica_jung_2017,ghica_compositional_2023} and, indeed in conventional e-graphs to encode infinite equivalence classes, as sketched in Fig.~\ref{fig:app2}. All these application domains stand to benefit from the use of optimisation techniques, which in turn can take advantage of the appropriate e-graph technique.

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{discrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{c,X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
        \\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
    % \begin{align*}
    % S_{E} &= \{\\
    % &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    % &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    % &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    % &\}
    % \end{align*}
and let relations $R_{V},R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

\begin{proposition}
\label{prop:complement_existence}
    Consider a pushout square in $\catname{EHyp_{\Sigma}}$ below
    % https://q.uiver.app/#q=WzAsNSxbMCwwLCJYIl0sWzIsMCwiWiJdLFsyLDIsIkMiXSxbMCwyLCJQXntcXHVyY29ybmVyfSJdLFsyLDUsIlxcYnVsbGV0Il0sWzEsMCwiaSIsMl0sWzEsMl0sWzIsM10sWzAsMywibSIsMl1d
% \[\begin{tikzcd}
% 	X && Z \\
% 	\\
% 	{P^{\urcorner}} && C \\
% 	\\
% 	\\
% 	&& \bullet
% 	\arrow["i"', from=1-3, to=1-1]
% 	\arrow[from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% 	\arrow["m"', from=1-1, to=3-1]
% \end{tikzcd}\]
\[
    \tikzfig{appendix/pushout_square}
\]
    Such that $m,f$ are monos and $\mathcal{K}$ is discrete.
    Then, pushout complement $\mathcal{K} \to \mathcal{L}^{\bot} \to \mathcal{G}$ exists if
    \begin{enumerate}
        \item $[m_{V}(f_{V}(v_i))) = [m_{V}(f_{V}(v_j)))$ for all $v_i,v_j$ in $V_{\mathcal{K}}$.
        \item $\consistency(m_{V}(f(v_i))) = \consistency(m_{V}(f_{V}(v_j)))$ for all $v_i,v_j$ in $V_{\mathcal{K}}$.
        \item The no-dangling condition holds.
    \end{enumerate}
\end{proposition}
\begin{proof}
    By requiring that $m$ is mono makes the no-identification condition satisfied on the nose.
    Then, no-dangling and no-identification conditions are necessary for pushout complement to exists~\cite{dpo}.
    Points (1)-(2) above guarantee that after removing from $\mathcal{G}$ everything in $\mathcal{L}$ that has no pre-image in $\mathcal{K}$ the arrow $\mathcal{K} \to \mathcal{L}^{\bot}$ will satisfy the assumptions~\ref{pushout:assumptions}.
    In particular, they mean that after removing the image of $\mathcal{L}$ from $\mathcal{K}$ we have the following
    \begin{itemize}
        \item $[c_{V}(v_{i})) = [c_{V}(v_{j}))$ for all $v_{i},v_{j}$ in $V_{\mathcal{K}}$.
        \item $\consistency(c_{V}(v_i)) = \consistency(c_{V}(v_j))$ for all $v_i,v_j$ in $V_{\mathcal{K}}$.
    \end{itemize}
    This means that $\mathcal{G}$ is isomorphic to the pushout constructed accroding to Theorem~\ref{th:existence_of_pushouts}.
\end{proof}

@@

\begin{figure*}[t!]
    \centering
    \[
        \scalebox{0.45}{
            \tikzfig{combinatorial_semantics/structural_rule_pushout}
        }
    \]
    \caption{The image of the top left equation from Figure~\ref{fig:string-equations} under $\llbracket - \rrbracket$}
    \label{fig:structural_rule_pushout}
\end{figure*}

@@

\begin{proposition}[Proposition~\ref{prop:soundness} restatement]
The category $\WellTypedMdaEcospans/{\mathcal{S}}$ is a semilattice-enriched PROP. 
\end{proposition}
\begin{proof}
    To prove this statement we essentially need to show that the functor $\llbracket - \rrbracket$ is well-defined, \textit{i.e.}, if $g = h$ in $\textsf{PROP}^{+}(\Sigma)$ then $\llbracket g \rrbracket = \llbracket h \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$ or $\llbracket g \rrbracket \Rrightarrow_{\mathcal{S}}^{*} \llbracket h \rrbracket$ in $\WellTypedMdaEcospans$.
    $g = h$ implies that there is a sequence of SMC and enrichment equations such that when applied to $g$ it is turned into $h$.
    As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \llbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[-]$ is defined by computing the pushout along the common interfaces.
    We first note that any subterm of $g$ correspond to a convex down-closed subgraph of $\llbracket f \rrbracket$ and hence there is always a convex down-closed match of $\llbracket l \rrbracket$ within $\llbracket g \rrbracket$.
    We then need to show that for all $\llbracket l \rrbracket = \llbracket r \rrbracket$ in $\mathcal{S}$ the match satisfies the conditions in Proposition~\ref{prop:complement_existence} which will complete the first square of EDPOI and then check that the arrows in the second square satisfy the condition from Theorem~\ref{th:existence_of_pushouts}.

    Consider the rule $\llbracket l \rrbracket = \llbracket r \rrbracket$ in Figure~\ref{fig:structural_rule_pushout} which corresponds to the top-left equation in Figure~\ref{fig:string-equations}.
    And recall the diagram assuming $\mathcal{L} = \llbracket l \rrbracket$, $\mathcal{R} = \llbracket r \rrbracket$, $\mathcal{G} = \llbracket g \rrbracket$ and $\mathcal{H} = \llbracket h \rrbracket$.
    \[
        \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square}
        }
    \]
    We then further note that if $l$ is a subterm of $g$ then the nodes in the image of input and output external interfaces of $\llbracket l \rrbracket$ in $\llbracket g \rrbracket$ all share the same set of parents and are pairwise consistent.
    This satisfies conditions (1)-(2) in Proposition~\ref{prop:complement_existence}.
    The no-dangling condition is satisfied because we consider only mda-cospans.
    Therefore, there exists a pushout complement $\mathcal{C}$ which is $\llbracket g \rrbracket$ with $\llbracket l \rrbracket$ removed.
    Then necessarily verticies in $\mathcal{C}$ in the image of $i+j$ share the same set of parents and are pairwise consistent as they do in $\mathcal{G}$ so.
    The same is true for the image of $i + j$ in $\mathcal{R}$ and therefore the necessary pushout exists which is $\mathcal{C}$ with $\mathcal{R} = \llbracket r \rrbracket$ glued in and it is equal to $\llbracket h \rrbracket$.
    Hence $\llbracket g \rrbracket \Rrightarrow{} \llbracket h \rrbracket$.
    The rest follows by induction on the number of applied equations.
    Similar argument then applies to other rules in $\mathcal{S}$ as well.
\end{proof}
